---
title: "RCode For Monte Carlo Least Squares"
output: pdf_document
date: "2025-04-17"
urlcolor: blue
---
# (Appendix A) Estimation of Parameters
[Estimation of Parameters Notebook Github](https://github.com/lamseamus/MAT4372FinalProject/blob/main/Estimate_Parameters.ipynb)

# (Appendix B) Binomial Model Code
## Function to price an American put using a Binomial Tree
```{r,eval=FALSE}
binomial_tree_american_put <- function(S0, K, T, r, sigma, M) {
  dt <- T / M  #Time Step Size 
  u <- exp(sigma * sqrt(dt)) # Upturn factor
  d <- exp(-sigma * sqrt(dt)) # Downward factor
  p <- (exp(r * dt) - d) / (u - d) # Risk-neutral probability

  # Stock price tree
  stock_tree <- matrix(0, nrow = M + 1, ncol = M + 1) # Create matrix to store stock prices
  for (j in 0:M) {
    for (i in 0:j) {
      stock_tree[i + 1, j + 1] <- S0 * u^i * d^(j - i) 
      # Number of steps multiplied by upturn and downturn factors
    }
  }
  # Option value tree
  option_tree <- matrix(0, nrow = M + 1, ncol = M + 1) # Matrix to store option values 
  option_tree[, M + 1] <- pmax(K - stock_tree[, M + 1], 0)  
  # Payoff is max(K - S, 0) at maturity, applied to entire column M+1

  # Work backward through the tree
  for (j in (M - 1):0) { # Go through each column in reverse order starting at M-1  
    for (i in 0:j) { # Go though each row in the column
      continuation <- exp(-r*dt)*(p*option_tree[i+2,j+2]+(1-p)*option_tree[i+1,j+2])
      exercise <- pmax(K - stock_tree[i + 1, j + 1], 0)
      option_tree[i + 1, j + 1] <- pmax(continuation, exercise)
    }
  }
  
  return(option_tree[1, 1])  # Option price at root
}

```

# (Appendix C) Monte Carlo Algorithm
## Function to price an American put using Monte Carlo simulations
```{r,eval=FALSE}
price_american_put_longstaff_schwartz_MC <- function(K, M, N, r, S0,sigma, polynomial) {
  
  dt <- 1/M
  discount <- exp(-r * dt)  
  set.seed(123)
  Z <- matrix(rnorm(N * M), nrow = N, ncol = M) # Vectorize Brownian Motion Simulation  
  S <- S0 * exp(sigma * sqrt(dt) * t(apply(Z, 1, cumsum)))
  
  Cash_flow <- matrix(0, nrow = N, ncol = M)
  Cash_flow[, M] <- pmax(K - S[, M], 0) 
  
  # Cash Flows at each time step 
  for (m in M:2) {
    X <- S[, m-1]
    Y <- Cash_flow[, m] * discount
    XY <- cbind(X, Y)
    XY[X > K, ] <- NA  

    if (all(is.na(XY))) {
        Cash_flow[, m-1] <- 0  # Skip regression if no in the money paths
        next
    }
    regression <- lm(polynomial, data = as.data.frame(XY))
    immediate_exercise <- pmax(K - S[, m-1], 0)
    continuation <- predict(regression, newdata = as.data.frame(X))
    Cash_flow[, m-1] <- ifelse(continuation < immediate_exercise, immediate_exercise, 0)
  }
  
  # Discounting cash flows
  for (i in 1:nrow(Cash_flow)) {
    for (j in 1:ncol(Cash_flow)) {
      if (Cash_flow[i, j] != 0) { 
        Cash_flow[i, j] <- Cash_flow[i, j] * round(exp(-r * j), 5)
        if (j < ncol(Cash_flow)) {
          Cash_flow[i, (j+1):ncol(Cash_flow)] <- 0
        }
        break
      }
    }
  }
  
  option_price <- mean(rowSums(Cash_flow))
  return(option_price)
}
```

# (Appendix D) Effect of Time Steps M 
[Effect of Time Steps M Notebook Github](https://github.com/lamseamus/MAT4372FinalProject/blob/main/Effect_Time_Steps.ipynb)

# (Appendix E) Monte Carlo Algorithm using Laguerre and Hermite polynomials 
## Function to price an American put using Laguerre polynomials
```{r,eval=FALSE}
price_american_put_Laguerre <- function(K, M, N, r, S0, sigma, degree) {
  
  dt <- 1/M
  discount <- exp(-r * dt)  
  set.seed(123)
  Z <- matrix(rnorm(N * M), nrow = N, ncol = M)
  S <- S0 * exp(sigma * sqrt(dt) * t(apply(Z, 1, cumsum)))
  
  Cash_flow <- matrix(0, nrow = N, ncol = M)
  Cash_flow[, M] <- pmax(K - S[, M], 0)
  
  laguerre_basis <- function(x, d) {
    L <- list()
    L[[1]] <- rep(1, length(x))                                  # L0
    if (d >= 1) L[[2]] <- -x + 1                                  # L1
    if (d >= 2) L[[3]] <- 0.5 * (x^2 - 4*x + 2)                   # L2
    if (d >= 3) L[[4]] <- ( -x^3 + 9*x^2 - 18*x + 6 ) / 6         # L3
    if (d >= 4) L[[5]] <- (x^4 - 16*x^3 + 72*x^2 - 96*x + 24) / 24 # L4
    if (d >= 5) L[[6]] <- (-x^5 + 25*x^4 - 200*x^3 + 600*x^2 - 600*x + 120) / 120  # L5
    do.call(cbind, L[1:(d+1)])
  }
  
  for (m in M:2) {
    X <- S[, m-1]
    Y <- Cash_flow[, m] * discount
    
    in_the_money <- X < K
    if (sum(in_the_money) == 0) {# Skip regression if no in the money paths
      Cash_flow[, m-1] <- 0
      next
    }
    
    X_in <- X[in_the_money]
    Y_in <- Y[in_the_money]
    
    laguerre_features <- laguerre_basis(X_in, degree) # Create matrix with laguerre degree 
    regression_df <- data.frame(Y = Y_in, laguerre_features)
    colnames(regression_df) <- c("Y", paste0("L", 0:degree))
    
    regression_formula <- as.formula(paste("Y ~", paste(colnames(regression_df)[-1], collapse = " + ")))
    regression <- lm(regression_formula, data = regression_df)
    
    all_features <- laguerre_basis(X, degree)
    colnames(all_features) <- paste0("L", 0:degree)
    continuation <- predict(regression, newdata = as.data.frame(all_features))
    
    immediate_exercise <- pmax(K - X, 0)
    Cash_flow[, m-1] <- ifelse(continuation < immediate_exercise, immediate_exercise, 0)
  }

  # Discounting cash flows
  for (i in 1:N) {
    for (j in 1:M) {
      if (Cash_flow[i, j] != 0) {
        Cash_flow[i, j] <- Cash_flow[i, j] * round(exp(-r * j), 5)
        if (j < M) Cash_flow[i, (j+1):M] <- 0
        break
      }
    }
  }

  return(mean(rowSums(Cash_flow)))
}
```

## Function to price an American put using Hermite Polynomials
```{r,eval=FALSE}
price_american_put_Hermite <- function(K, M, N, r, S0, sigma, degree) {
  
  dt <- 1/M
  discount <- exp(-r * dt)  
  set.seed(123)
  Z <- matrix(rnorm(N * M), nrow = N, ncol = M)
  S <- S0 * exp(sigma * sqrt(dt) * t(apply(Z, 1, cumsum)))
  
  Cash_flow <- matrix(0, nrow = N, ncol = M)
  Cash_flow[, M] <- pmax(K - S[, M], 0)
  
  hermite_basis <- function(x, d) {
    H <- list()
    H[[1]] <- rep(1, length(x))                                  # H0
    if (d >= 1) H[[2]] <- x                                       # H1
    if (d >= 2) H[[3]] <- x^2 - 1                                 # H2
    if (d >= 3) H[[4]] <- x^3 - 3*x                               # H3
    if (d >= 4) H[[5]] <- x^4 - 6*x^2 + 3                        # H4
    if (d >= 5) H[[6]] <- x^5 - 10*x^3 + 15*x                    # H5
    do.call(cbind, H[1:(d+1)])
  }
  
  for (m in M:2) {
    X <- S[, m-1]
    Y <- Cash_flow[, m] * discount
    
    in_the_money <- X < K
    if (sum(in_the_money) == 0) {
      Cash_flow[, m-1] <- 0
      next
    }
    
    X_in <- X[in_the_money]
    Y_in <- Y[in_the_money]
    
    hermite_features <- hermite_basis(X_in, degree)# Create matrix with Hermite degree 
    regression_df <- data.frame(Y = Y_in, hermite_features)
    colnames(regression_df) <- c("Y", paste0("H", 0:degree))
    
    regression_formula <- as.formula(paste("Y ~", paste(colnames(regression_df)[-1], collapse = " + ")))
    regression <- lm(regression_formula, data = regression_df)
    
    all_features <- hermite_basis(X, degree)
    colnames(all_features) <- paste0("H", 0:degree)
    continuation <- predict(regression, newdata = as.data.frame(all_features))
    
    immediate_exercise <- pmax(K - X, 0)
    Cash_flow[, m-1] <- ifelse(continuation < immediate_exercise, immediate_exercise, 0)
  }

  # Discounting cash flows
  for (i in 1:N) {
    for (j in 1:M) {
      if (Cash_flow[i, j] != 0) {
        Cash_flow[i, j] <- Cash_flow[i, j] * round(exp(-r * j), 5)
        if (j < M) Cash_flow[i, (j+1):M] <- 0
        break
      }
    }
  }

  return(mean(rowSums(Cash_flow)))
}
```

# (Appendix F) Convergence plots for Regular, Laguerre, and Hermite Polynomials
[Convergence plots for Regular, Laguerre, and Hermite Polynomials Notebook Github ](https://github.com/lamseamus/MAT4372FinalProject/blob/main/PlotsforPolynomials.ipynb)

# (Appendix G) Monte Carlo Algorithm using Custom Polynomials 
## Black-Scholes formula for European call and put options 
```{r,eval=FALSE}
# Black-Scholes formula for European call
bs_call <- function(S, K, r, T, sigma) {
  set.seed(123)
  d1 <- (log(S / K) + (r + 0.5 * sigma^2) * T) / (sigma * sqrt(T))
  d2 <- d1 - sigma * sqrt(T)
  S * pnorm(d1) - K * exp(-r * T) * pnorm(d2)
}

# Black-Scholes formula for European put
bs_put <- function(S, K, r, T, sigma) {
  set.seed(123)
  d1 <- (log(S / K) + (r + 0.5 * sigma^2) * T) / (sigma * sqrt(T))
  d2 <- d1 - sigma * sqrt(T)
  K * exp(-r * T) * pnorm(-d2) - S * pnorm(-d1)
}
```

## Function to price an American put using European call and put options
```{r,eval=FALSE}
price_american_put_longstaff_schwartz_MC_euro <- function(K, M, N, r, S0, sigma, polynomial) {
  dt <- 1 / M
  discount <- exp(-r * dt)  
  set.seed(123)
  Z <- matrix(rnorm(N * M), nrow = N, ncol = M)  
  S <- S0 * exp(sigma * sqrt(dt) * t(apply(Z, 1, cumsum)))
  
  Cash_flow <- matrix(0, nrow = N, ncol = M)
  Cash_flow[, M] <- pmax(K - S[, M], 0)
  
  for (m in M:2) {
    X <- S[, m - 1]
    T_remaining <- (M - m + 1) * dt

    call_bs <- bs_call(X, K, r, T_remaining, sigma)
    put_bs <- bs_put(X, K, r, T_remaining, sigma)

    df_reg <- data.frame( # Create regressors using custom functions 
      S = X,
      call = call_bs,
      put = put_bs,
      Y = Cash_flow[, m] * discount
    )

    df_reg[X > K, ] <- NA

    if (all(is.na(df_reg))) {
      Cash_flow[, m - 1] <- 0
      next
    }

    regression <- lm(polynomial, data = df_reg)
    immediate_exercise <- pmax(K - X, 0)

    df_pred <- data.frame(
      S = X,
      call = call_bs,
      put = put_bs
    )
    continuation <- predict(regression, newdata = df_pred)
    Cash_flow[, m-1] <- ifelse(continuation < immediate_exercise, immediate_exercise, 0)
  }
  
  # Discounting cash flows
  for (i in 1:nrow(Cash_flow)) {
    for (j in 1:ncol(Cash_flow)) {
      if (Cash_flow[i, j] != 0) {
        Cash_flow[i, j] <- Cash_flow[i, j] * round(exp(-r * j * dt), 5)
        if (j < ncol(Cash_flow)) {
          Cash_flow[i, (j + 1):ncol(Cash_flow)] <- 0
        }
        break
      }
    }
  }

  return(mean(rowSums(Cash_flow)))
}
```


# (Appendix H) Covergence Plots using European Options as Regressors 
[Convergence Plots using European Options as Regressors Notebook Github  ](https://github.com/lamseamus/MAT4372FinalProject/blob/main/PlotsforEuropeanRegressors.ipynb)


# (Appendix I) Convergence Error with Degree In-The-Money
[Convergence Error with Degree In-The-Money Notebook Github ](https://github.com/lamseamus/MAT4372FinalProject/blob/main/ConvergenceErrorDegreeInMoney.ipynb)



































